---
title: "Opinion Change Modeling and Measurement"
date: 2025-09-24
author: Matt Turner
categories: [opinion dynamics, polarization, extremism, statistics]
number-sections: false
editor: 
  markdown: 
    wrap: 80
---

```{r}
#| include: false
library(tidyverse)
library(tidytext)

```

**Abstract**

Social influence shapes opinions in ways that matter for sustainability. Unlike behaviors, which are adopted or not, opinions shift gradually—toward, away from, or resistant to those of others. We can think of assimilation as a scale: at one end, consensus, where people grow more similar through interaction; at the other, anti-consensus, where disagreement deepens; and in the middle, a zone of zero influence, through ignorance or defiance. These dynamics aren’t visible in behavior alone. A closer look at group polarization shows how extreme, inflexible actors can shift group outcomes even when average opinions appear unchanged. If we understand how opinions move—not just what people do—we can design influence, not just react to it.


# Introduction: opinion change to reduce cultural inertia against sustainability

Extremism and polarization can incapacitate the governments and institutions who are responsible for ecological and environmental protection. Opinions are equivalent to beliefs or attitudes: they are theoretical constructions hypothesized to affect behavior, but need not be based on evidence or experience, but hearsay from others. 
In forest management, we know prescribed burns could prevent catastrophic wildfires
[@Eisenberg2019; @Kolden2019]. So although we have Indigenous practices for
climate action ([SDG 13](https://sdgs.un.org/goals/goal13)) and the
protection of Life on Land ([SDG 15](https://sdgs.un.org/goals/goal15)) they
are ignored in favor of maladaptive modern practices. These traditional
practices have helped historically local inhabitants over thousands of
years, and they can still help now. 

This is not to say that the healthiest institutions have zero differences in opinion among its members, or that all citizens be impassionate moderates on every issue. Healthy disagreement is necessary to be able to adapt to the steady march of history punctuated by acute crises—times when cultural diversity within an institution is harnessed as a source of innovation to adapt and persist. Furthermore, I, for one, am 100% FOR *ELEVATED* (not "extreme", see?) support for policies that enforce agreements to protect our collective rights to life, liberty, and the pursuit of happines, and to protect these same rights for future generations[^ethics-of-enforcement]. 

The science of opinion change, or just *opinion science*, has a critical role in addressing the puzzle of viral sustainability—what will it take for more simple, inexpensive, proven methods for climate action and ecological protection [@fig-puzzle-of-adaptation]? In some cases, extremism turns violent and destructive, endangering hope for our collective future, sustainable or not. Tensions flare when polarization increases ever high—interested parties can rapidly tip over from uneasy coexistence to open violence. 

Diversity and passion can fuel institutions perpetually so that "failure is impossible"—these were the rousing final words of Susan B. Anthony to the women's suffrage movement she led. The "extremism" of those demanding women's suffrage was necessary for the moment, since the choice is binary—just as it is now for designing sustainable futures. The alliance between women, black, and universal suffrage movements combined to be more than the sum of its parts because of the way diverse voices amplify one another, and the gain of the amplification increases exponentially with the number and diversity of the collective.

In the following notes we review first how opinions and opinion change are measured. I'll give real-world examples from studies of *group polarization*, which is defined as when people in echo chambers become more extreme in their views over time just by talking with one another.  Then I will introduce a model of social influence that drives opinion change. Opinion change also depends on personality types in a population and patterns in interactions between people. Some people interact more frequently than others. We use *social network models* to represent different interaction structures.

![**The puzzle of
adaptation.**](../Figures/puzzle-of-adaptation.png){#fig-puzzle-of-adaptation
width="700"}



# Measurement



Opinion dynamics are commonly studied empirically using 
survey studies administered to the same group of participants over
time, called *longitudinal* surveys. We will focus on this interpretation for
  concreteness and because it is the most used measurement method in group polarization studies, which I use for a case study below. Unfortunately, group polarization presents a textbook case of what *not* to do to measure opinion change—it seems statistical problems have caused some 
  
## Survey measurement 



## Implicit Association Test

- Other options for measuring opinions:
  1. *implicit association test*
   uses keypress timing in an computer-based identification task to
  infer opinions, or *bias*, in 
    - see @Greenwald2009 for a more recent analysis, @Greenwald1995 for one
      of the landmark original papers
    - For more research and teaching materials, the [Project Implicit
      Resources]() looks impressive, including: 
      - [About the
        IAT](https://www.projectimplicit.net//resources/about-the-iat/), which
        explains it very well[^1] 
      - [Stimulus Materials for Social and Behavioral
      Research](https://www.projectimplicit.net/resources/study-materials/).
      - [Published manuscripts using **Public Data**](). To me this could be a
        treasure trove for project ideas. Think about modifying one or mashing
        up a couple of these and it seems you could have an impactful project.
        

## Computational linguistics: Words as Windows into Opinion and Meaning

*Computational linguistics* provides several tools for inferring peoples' opinions, or what different words mean to different communities over time, that could be useful for promoting sustainable futures. I breifly mention two below with some references and examples: *sentiment analysis* that infers a person's opinion (equivalent to sentiment I say) and *topic analysis* that calcualtes the probabilty one word tells us about the meaning of another depending on how frequently, they occur with each other compared to other words.


### Sentiment Analysis

*Sentiment analysis* is a computational linguistic method
that infers a writer or speaker's opinion on different topics based on how
frequently they used positive or negative words (such as "good" or "bad") 
in discussing those topics. Surveys elicit what people are ready to declare; the IAT stages associations to catch what cannot be comfortably admitted. A third approach is to look outward, treating ordinary language—how people talk about energy, climate, or community—as opinion in itself [@Cody2015]. 

Sentiment analysis makes this tractable at scale: tokenize text, map words to a sentiment lexicon, and tally the balance. In R this takes only a few lines, and while the results are blunt (“hope” = positive, “fear” = negative), bluntness is revealing. These tags have no essence: hope is only hope against despair, good against bad, and in other contexts fear preserves, hope deceives, evil sustains. Words are not timeless carriers of meaning but cultural agreements we stabilize for a time. 

What sentiment analysis exposes, then, is not hidden psychological states but the scaffolding by which communities themselves domesticate affects into usable binaries. For sustainability research this is crucial: it shows how opinion change unfolds not in pristine categories but in the contingent ways we decide to commit to certain meanings and let others go. If we try to be everything to everyone we will be nothing in the end.


```{r}
text <- tibble(
  id = 1:3,
  message = c(
    "Solar energy adoption is inspiring and creates hope for a sustainable future.",
    "Climate change is a serious threat that causes anxiety and fear.",
    "Community gardens reduce waste and bring joy to neighborhoods."
  )
)

tidy_text <- text %>% unnest_tokens(word, message)

sentiments <- tidy_text %>%
  inner_join(get_sentiments("bing"), by = "word")

sentiment_summary <- sentiments %>%
  count(id, sentiment) %>%
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) %>%
  mutate(overall_sentiment = positive - negative)
```

### Topic Modeling

*Topic Modeling* is a similar technique that *does* help us infer word meanings when said by different people in different contexts over time [@Hamilton2016;@Hamilton2016c]. The technique tracks
how frequently words co-occur with each other in phrases and sentences to quantify how much more likely they are than chance to be part of one another's conceptualization—for example, .

The example below is not very interesting because our *corpus*, or the set of "documents" that we analyze, is too small to perform the statistics necessary to learn the meanings of words.

For politics, decision making, or sustainability it could be helpful to explore how terms like *sustainability*, *climate change*, and *adaptation* are used on different news outlets, online communities, and public forums, just to name a few opportunities.

```{r}
library(tidyverse)
library(tidytext)
library(topicmodels)


# Tokenize and create document-term matrix
tidy_text <- text %>%
  unnest_tokens(word, message) %>%
  anti_join(stop_words, by = "word")

dtm <- tidy_text %>%
  count(id, word) %>%
  cast_dtm(document = id, term = word, value = n)

# Fit a 2-topic LDA model
lda_model <- LDA(dtm, k = 2, control = list(seed = 123))

# Get per-document topic probabilities (gamma)
doc_topics <- tidy(lda_model, matrix = "gamma")

print(doc_topics)
```

# Opinion Dynamics Models

The scientific study of change in the physical world is the study of *dynamics*.
So, the scientific study of opinion change is *opinion dynamics*,
a label shared by sociologists, cognitive scientists, and species of 
social scientist. In our simulations, we observe several outcomes: whether consensus or polarization occur either globally or in certain communities; and whether a group's opinion becomes more extreme when it finds consensus, or whether the magnitude of the 
mean group opinion stays constant or increases in magnitude. If it increases in magnitude this is indicative of *group polarization*, a central research topic in social psychology.

Opinion dynamics can take several forms, but we can trace a through-line of
the development of what we can now call *social influence* opinion dynamics.
Social influence dynamics calculate the *influence* that one person's opinions
have on another person's, with discourse or discussion being the "force carrier" 
through which influence is achieved. The *social influence* on a person's
opinions is a scalar value, which we denote $\Delta o$, where 
$\Delta$ is the classic prefix we add to mean "change in" the variable that
follows it—the opinion $o$, in this case. The sign of one's opinion $o$ 
represents the *valence* of their opinion—how "negative" or "positive". 
The magnitude of one's opinion represents their extremism [@Blau1974]. Note this
implies *opinions* exist within limited affordances of different *topics*, 
which themselves are latent and unobservable.



### Social Influence

Social *influence* can change opinions a little or a lot. This is different
  from the social *learning* that fuels behavior change, where someone either
  does something or they don't—it's binary.
In social influence, people can either develop similar opinions as their
peers, or ignore their peers' opinions, just as people can either adopt or
ignore their peers' *behaviors* in social learning. In the language of
opinion dynamics, peers find *consensus*, subject to *biased assimilation*.

*Consensus* means a group of two or more people's opinions become more similar over time. *Anti-consensus* is the opposite, where when people discuss their opinions, their opinions become more different—it is like the opinions *repel* each other. Consensus, then, is identified by a reduction in variance of group opinions, and anti-consensus by an increase in variance. Consensus, then, leads to a reduction in polarization, and anti-consensus an increase in polarization. 

When consensus occurs the mean opinion could stay constant or change along with the reduction in variance. If the mean stays constant this is *simple consensus*. If the mean magnitude of group opinions becomes greater over time, then the group became more extreme, otherwise the group became more moderate.


### A Very Brief and Incomplete History

@Abelson1964 perhaps first sketched a model like the one we adopt here. His chapter reviews mathematical approaches to opinion change of the time, including terms that modulate the strength of influence such as opinion difference and opinion extremism.
@DeGroot1974 developed an early *simple consensus* influence dynamic where 
interaction always increases opinion similarity through iterated averaging.
The *bounded confidence* dynamic developed a mechanism of selective ignorance activated if people's opinions were sufficintly outside some "bound"—which inspires "confidence" I guess [@Hegselmann2002;@Lorenz2007]?


### Social Networks and Initial Conditions


@Macy2003 first introduced what I call *social influence* dynamics. 
Social influence dynamics incorporates gradual consensus or anti-consensus—
in other words, opinion "repulsion"—plus the tendency to *biased
assimilation* of information, where people ignore information that is
neither supportive or challenging enough to catch their interest. This
model was adapted from the Hopfield network model to explain the physical processes
underlying memory in cognitive science (see margin note right). 

@Flache2011 took this a step further to investigate how social networks
might affect whether societies became highly polarized or not. They
modeled how the internet increases our exposure to opinions 
we are likely to hate, leading to increased
polarization—that's ignoring any algorithmic boosts for such content. 
They found that adding such "long-range ties" between previously
disconncted people increased opinion polarization in simulations.

In @Turner2018 we adopted a similar approach to understand the complex causal mechanisms 
show that the timing of different 
interactions, and how well we communicate, of human discourse, a society's initial
extremism, and shared cultural interest
There are many unanswered questions about how different structures of
interaction, which we represent in computer simulations as social network models.
This may be especially salient in the case of either de-escalation campaigns
to reduce a society's more destructive tendencies, or in trying to 
strengthen resolve and conviction to pursue sustainable policies.


::: {.column-margin}
**Stubborn extremism: a hidden assumption of the Hopfield model.** 
@Hopfield1982 first developed this model to simulate 
physical mechanisms for memory in
analog-valued neural networks. Neurons have a maximum amount they can be
electro-chemically activated either positively or negatively while they do their job in our brains.
Hopfield encoded this in his model as a smoothing factor that made neuron
activation
change more slowly when closer to ±1. In social influence, this limitation of 
neurons becomes "stubborn extremism" automatically. Stubborn extremism has relatively rich
empirical support, which I review in @Turner2020—but its original inclusion
in social influence dynamics appears to be a coincidence!
:::

These approaches can generate emergent social phenomena like radicalization
(*group polarization*) from empirically-motivated model assumptions about social
influence dynamics. This signals to us that we have a plausible mechanistic
model: a causal story supported by evidence about real-world social influence
mechanisms produced by a pluralism of social science disciplines. 

### Social influence formalism

Each individual, $i$, has a set of $C$ opinions with a vector of latent opinions 

$$
\mathbf{o}_i = (o_{i1}, o_{i2}, \ldots, o_{iC}),
$$ {#eq-opinion-vector}

where $C$ is the cultural complexity (number of opinion dimensions).  


### Stubbornness

Stubbornness scales the responsiveness of an opinion to social influence:

$$
s(o;\alpha) \;=\; \frac{1}{1 + |o|^\alpha}.
$$ {#eq-stubbornness}

Moderate opinions ($o \approx 0$) are highly flexible ($s(o)\approx 1$), while
extreme opinions ($|o| \gg 0$) become increasingly stubborn ($s(o) \to 0$).
This form works for both bounded and unbounded opinions: bounded models impose
limits externally, while unbounded models retain smooth decay.

### Dyadic Interaction

When focal agent $i$ interacts with partner $j$, each opinion component is
updated by

$$
o_{i,c}' \;=\; o_{i,c} + \alpha \Delta o_{i,c},
$$ {#eq-dyadic-update}

where $c \in \{1, \dots, C\}$ indexes the cultural dimension—the topic the opinion
refers to, $\alpha \in [0,1]$ is the global learning rate, and

$$
\Delta o_{i,c} = \alpha \, s(o_{i,c}) \,\bigl(o_{j,c} - o_{i,c}\bigr)
$$


### Stubborn extremism

- $s(o_{i,c})$ is stubbornness (@eq-stubbornness).


### Iteration

After all dyadic interactions in a time step, opinions are updated synchronously:

$$
\mathbf{o}_i \;\leftarrow\; \mathbf{o}_i'.
$$ {#eq-iteration}
  

# Case study: echo chamber radicalization (aka *group polarization*)

-   We can use the models and measurement we learned so far to
    1.  simulate *echo chamber radicalization*, which we will call *group
        polarization* because it's shorter and that's it's name for social
        psychologists. But we tend to think of polarization as being when two
        opposing groups are extremely rigidly opposed to each other; some
        sociologists call this "bi-polarization" to differentiate the everyday
        layperson's definition of "polarization" from what social psychologists
        called "group polarization".
    2.  show how we can estimate the false discovery rate and statistical power
        of hone experimental design to calculate the false discovery rate


# Discussion

- We introduced methods for measuring real-world opinion change and simulating latent opinion dynamics and the survey measurement process.
- How we represent opinion change has practical implications: we saw that a high number of select group polarization studies must be retracted because the detections of group polarization they present do not exclude the possibility that simple consensus and not group polarization occurred. Therefore the findings must be excluded from our considered research.
- Opinion dynamics is founded on a long interdisciplinary history, which we plan to continue to build on.
- There are immediate research opportunities because we know little about real-world opinion dynamics in general—our lack of understanding is worse still in sustainability sciences.
